---
title: "Environmental and genetic contributions to eco-geographic rules in house mice"
author: "Mallory A. Ballinger"
date: "12/19/2020"
output:
  github_document:
    html_preview: false
---
[Abstract goes here]
Using metadata from museums, colony data from the Nachman lab, and morphological data generated from a common garden experiment, we examine the environmental and genetic contributions of Bergmann's Rule and Allen's Rule in temperate and tropical populations of house mice.


### Overview:
--------

    BallingerNachman_allenbergmann_XXXX_2021
    |- README          # top level description of content
    |
    |- data            # raw and primary data, are not changed once created
    |  |- raw/         # raw data, will not be altered
    |  +- process/     # cleaned data, will not be altered once created
    |
    |- code/           # any programmatic code
    |
    |- results/        # all output from workflows and analyses
    |  |- tables/      # tables, likely designated for manuscript tables
    |  |- figures/     # graphs, likely designated for manuscript figures
    |
    |- submission      # files for manuscript
    |
    +- Makefile        # executable Makefile for this study

```{r libraries, message=FALSE, echo=FALSE}
library(tidyverse)
library(rmarkdown)
library(here)
```

### Dependencies:  
* `r R.version.string`  
    * `tidyverse` (v. `r packageVersion("tidyverse")`)  
    * `rmarkdown` (v. `r packageVersion("rmarkdown")`)  
    * `here` (v. `r packageVersion("here")`)
    * `tinytex` (v. `r packageVersion("tinytex")`)


### My computer
```{r sessionInfo, echo=FALSE}
sessionInfo()
```